Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-03-30 jss/EpiCompare-paper/EpiCompare-paper.Rmd
We are moving right along.  Continued editing Section 2.

\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking a global, not marginal view of how epidemics move through populations and (2) by treating full epidemics as filamental trajectories^[\ben{[Ben says: Do we need a footnote explaining what this is more clearly?]}] and not points produced by functions of time\footnote{\shannon{[Shannon says: this notation needs to be cleaned up.]}} (this will be explained in Section \ref{sec:r0}). The following section aims to highlight the strengths of \textit{time-invariance} and define the mathematical foundations that \pkg{EpiCompare}’s tools stand upon.

Mathematically, epidemics are complex objects.  They can be hard to assess and compare to one another due to the differences in the diseases, the location where the outbreak occurs,  how the affected population reacts, and the time related features (including start of the epidemic, speed of infection and more). Time-invariant analysis make different epidemics easier to compare by removing many time dependent aspects of an epidemic. Instead it focuses on the overall trajectory of an epidemic, emphasizing the number of lives affected.

## Motivating time-invariance through $R_0$^[\ben{Is this still a relevant title for this section? Think about this header after the rewrite.}] {short-title="motivating through R0" #r0}

Time-invariance analysis, as it appears in \pkg{EpiCompare}, \sout{can be seen as an attempt to escape} \shannon{bypasses} many difficulties \shannon{in} comparing different epidemics. With time-invariant analysis, comparing the decades-long outbreak of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is \sout{still} possible. Time-dependent problems can arise when estimating epidemiological parameters, including the reproduction number $R_0$. \ben{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}

@@

$R_0$ is probably the most famous time-invariant numerical summary of an epidemic and is often associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2001}\footnote{[Shannon: correct CITation]}.  $R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a completely susceptible population \citep{anderson1992}. $R_0$ can be a difficult quantity to estimate and is typically sensitive to time-based parameters such as the beginning and end of an epidemic \citep{Gallagher2020}. \ben{[Ben says: \@Shannon could we expand on this difficulty / show it really is difficult with 1 more sentence? It might benefit from listing the difficulting in estimating a robust $R_0$ value (difficulties: time steps used, scale and "start" and final start, assumptions about population growth and underlying model fit).]} To demonstrate the difficulty of discerning $R_0$ in a time-dependent analysis, we first introduce the \citet{Kermack1927}’s SIR model. This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,

@@

## \ben{Time invariant visualization of differences in $R_0}\sout{ Visualizing $R_0$} {short-title="visualizing r0" #vis-r0}

\ben{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$? Also, I feel like the story is weak here. The point is to leverage $R_0$ to show the value of time-invariant analysis - this seems a bit more like just discussing properties of $R_0$. In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}

\ben{[Ben says: replacement of following paragraph: presentation rearranged to emphasis the figure first (when presenting the problem), and then discussing it.]} Traditional $state$ vs. $time$ plots, like seen in Fig. \ref{fig:different-scales-standard}, present information about epidemics in a time-dependent and piece-wise fashion, and make comparing different epidemic $R_0$s impossible. In Fig. \ref{fig:different-scales-standard} we compare [two different epidemics that start in two populations of 1000 people with 10 individuals initially infected, but with different infection and recovery rates (see Eq. \ref{eq:sir-ode})]. More specifically, we've observed both epidemics for 15 days, with the first epidemic\footnote{In the plot it says model - probably want to say epidemic?} (in red) having parameters $\beta_1, \gamma_1 = (1,2)$ and the second epidemic (in blue) having  parameters $\beta_2, \gamma_2 = (2,4)$. [At a glance, we may believe that \sout{Model}\ben{epidemic} 1 has a larger $R_0$ than \sout{Model}\ben{epidemic} 2 because the peak of infection occurs more quickly than in Model 2 \ben{(remember $R_0$ relates to the rate of infection)}.  On the other hand, we may believe \sout{Model}\ben{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \sout{model}\ben{epidemic} has not yet peaked at time 15. Yet, both epidemics have the same $R_0$ of 2].

\ben{[Ben says: bring Fig. \ref{fig:different-scales-standard} here.]}

\ben{[Ben says: this paragraph is replaced with the one above]}With regards to traditional epidemic $state$ vs. $time$ plots, when \sout{comparing}\shannon{examining} different epidemics their $R_0$s are often impossible to compare. This can be seen in a basic example, where we have two different epidemics that start in two populations of 1000 people with 10 individuals \shannon{initially} infected, but with \sout{different parameters} \shannon{different infection and recovery rates (see Eq. \ref{eq:sir-ode}}). Specifically suppose that $\beta_1, \gamma_1 = (a,b)$ and $\beta_2, \gamma_2 = (ca,cb)$ where $a, b, c > 0$. Further suppose we observe both of these epidemics for 15 days. The epidemic trajectories are shown in the $state$ vs. $time$ plots in Figure \ref{fig:different-scales-standard}.  At a glance, we may \sout{assume} \shannon{believe} that Model 1 has a larger $R_0$ than Model 2 because the peak of infection occurs more quickly than in Model 2.  On the other hand, we may \sout{think} \shannon{believe} Model 2 has a larger $R_0$ because we may think the number of infections in that model has not yet peaked at time 15. Yet, both epidemics have the same $R_0$ of 2.

\shannon{Introduce ternary plots here.  The below paragraphs are mostly new}

\ben{[Ben says: another rewrite of the paragraph below.]} A time-invariant approach to visualizing epidemics, which will be described more below, provides a simple way to compare $R_0$ values of epidemics that are generated from the Kermack and McKendrick SIR model. In these visuals, we treat epidemics as \textit{filamental trajectories}. A filamental trajectory can be mathematically viewed as a set of points is space that have an ordering, and that all points on the line between these order points are also contained in the geometric object. For a SIR epidemic, we can represent the associated filamental trajectory $\psi$ as

$$
\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
$$
where the an mapping $\xi : t \to \mathbb{R}$ that is monotonically increasing would not change the definition of $\psi$, i.e. $\psi_\xi = \psi$ where :

$$
\psi_\xi = \left \{(S(\xi(t)), I(\xi(t)), R(\xi(t))): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]}.
$$
[Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then \ben{all points in} $\psi$ will lay in a \ben{\sout{2d triangular plane, i.e. a ternary plot}} \ben{two-dimensional triangular plane in the three-dimensional space, which can be visualized in a two-dimensional ternary plot}.  As a result, we can visualize the global trajectory of an epidemic in a single \ben{ternary plot \sout{2d plot and ultimately $R_0$}}.  \shannon{[Shannon says: Show pic here?]}]\footnote{Which?}

\ben{[Ben says: this paragraph was replaced with the paragraph above]} \sout{However, time-invariant centric analysis allows us to visualize $R_0$ for the Kermack and McKendrick SIR model.  We call the epidemic trajectory of number of individuals in each state a \textit{filament}.  Mathematically, a filament $\psi$ for the SIR model  $\psi$ this is the set of points of the number in each state for each time point,}
$$
\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t=0:T}.
$$

\sout{Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then $\psi$ will lay in a 2d triangular plane, i.e. a ternary plot.  As a result, we can visualize the global trajectory of an epidemic in a single 2d plot and ultimately $R_0$.  \shannon{Show pic here?}}

\ben{[This section could be a bit less wordy. But generally good.]} We visualize the two epidemics in a global, ternary view in Figure \ref{fig:different-scales-tern}.  Without getting into too much detail of the intricacies of this plot, we immediately see the points of the two filaments $\psi$ seem to form the same trajectory.  Now, it is much clearer that \ben{\sout{Model} epidemic} 2 is following the same trajectory as \ben{\sout{Model} epidemic} 1 but is not as far along in the infection process.  We can see there is something fundamentally linking these two different epidemics, and this fundamental link turns out to be $R_0$.

We can show this mathematically\ben{.\sout{ if we} L}et our two epidemics be presented as $\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}$, $\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}$ respectively.\footnote{\ben{[Ben says: Originally this was a sentence fragment.]}}  As with the example, assume both models have the same initial values $(S(0), I(0), R(0))$, and let $R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$. And define $a>0$ to be the relative scalar such that $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$.

@@

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.  The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model \ben{\sout{is also} can be mapped to} a point on the epidemic path of the second SIR model.\ben{[We've changed how we describe these two objects when they are related to the theorem - is that desirable?]}

@@

```{r echo = FALSE, fig.cap = "\\label{fig:different-scales-standard}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$."  }
set.seed(1225)
theme_set(theme_bw(base_size = 24))

beta1 <- .8
gamma1 <- .4
scale <- .8
sir1 <- simulate_SIR_agents(n_sims = 100, n_time_steps = 40,
                            beta = beta1, gamma = gamma1,
                            init_SIR = c(990, 10, 0)) %>%
  group_by(sim) %>%
  agents_to_aggregate(states = c(tI, tR)) %>%
  group_by(t) %>%
  summarize(S = mean(X0), I = mean(X1), R = mean(X2),
            .groups = "drop") %>%
  mutate(type = "Model 1")

sir2 <- simulate_SIR_agents(n_sims = 100, n_time_steps = 40,
                            beta = beta1 * scale, gamma = gamma1 * scale,
                            init_SIR = c(990, 10, 0)) %>%
  group_by(sim) %>%
    agents_to_aggregate(states = c(tI, tR)) %>%
  group_by(t) %>%
  summarize(S = mean(X0), I = mean(X1), R = mean(X2),
            .groups = "drop") %>%
  mutate(type = "Model 2")

@@

Through the $R_0$ example, we see that treating epidemics like a filamental trajectories embedded in a lower dimensional space allows us to better compare the overall structure of the epidemic and see how the population was directly impacted. \ben{In this section we will present time-invariant tools that can be useful even when the underlying generative model for the epidemic is unknown. We will also present how this tools can be used when the epidemics have more than 3 states.} \sout{trajectories, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states. }

\ben{New paragraph} Viewing epidemics as filamental trajectories provide a lot new ways to compare and examine epidemics in a time-invariant manner. For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid impacts of temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point. This also creates a natural distance between this type of representation between epidemics, specifically:
\[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\psi_1'(s) - \psi_2'(s))^2 ds
\]
where $\psi_i'(s)$ the point along $\psi_1$ that is $s$ fraction of $|\psi_1|$ distance away from the start of $\psi_i$. This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall extremeness of points, including psuedo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}. This can be very useful when comparing between a set of simulation epidemics and the true epidemic, and does not constrain the number of states of the model, though we recommend projecting the points into the unit simplex\footnote{\ben{I'm not sure we've talked about this before...}} (by making all values the proportion of the population in the given state).

\ben{New paragraph:} If the set of epidemics that one is examining have only gone through a single cycle of the outbreak (before the population of individuals have become susceptible again), then additional time-invariant tools, including prediction regions can we leveraged. In these setting we go a step farther and treat epidemics more like geometric filaments than filamental trajectories. In \pkg{EpiCompare} we create prediction regions that  contain a the top ($1-\alpha$) proportion of simulated curves by defining geometric regions defined by the union of small filaments around the subset of simulations (subset by measures like the above psuedo-density estimates or depth estimates). These regions look at where in the state-space we expect the epidemic to traverse, and we can compare prediction regions defined by different models using the Haussdorf distance as well as examining how well the truth epidemic matches the simulations by examining if the epidemic's trajectory lies within the prediction region. All these geometrics structures and distances notations apply to epidemics with any number of states, and at the end of  Section \ref{sec:overview} we also highlight how these prediction regions can aid in visual comparisons for epidemics with 3 states (like the SIR models).

@@

>>>>

\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking a global, not marginal view of how epidemics move through populations and (2) by treating full epidemics as filamental trajectories and not points produced by functions of time \shannon{(these concepts will be explained in  explained in Section \ref{sec:r0})}. The following section aims to highlight the strengths of \textit{time-invariant} \shannon{analysis} and define the mathematical foundations that \pkg{EpiCompare}’s tools stand upon.

Mathematically, epidemics are complex objects.  They can be hard to assess and compare to one another due to the differences in the diseases, the location where the outbreak occurs,  how the affected population reacts, and the time related features (including start of the epidemic, speed of infection and more). Time-invariant analysis makes different epidemics easier to compare by removing many time dependent aspects of an epidemic. Instead, \shannon{time-invariant analysis} focuses on the overall \shannon{shape and direction, a filamental trajctory,} \sout{trajectory of }an epidemic.  \shannon{These filamental trajectories,  in turn,} emphasize the number of lives affected.

## Motivating time-invariant analysis through \shannon{the reproduction number} $R_0$^[\ben{Is this still a relevant title for this section? Think about this header after the rewrite.}] {short-title="motivating through R0" #sec:r0}

Time-invariant analysis, as it appears in \pkg{EpiCompare}, \shannon{bypasses} many difficulties \shannon{in} comparing different epidemics. With time-invariant analysis, comparing the decades-long outbreak of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is \sout{still} possible. Time-dependent problems can arise when estimating epidemiological parameters, including the reproduction number $R_0$. \sout{\ben{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}^[\shannon{I don't think this is a necessary sentence}]

@@

$R_0$ is probably the most famous time-invariant numerical summary of an epidemic and is often associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.  $R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a completely susceptible population \citep{anderson1992}.  \shannon{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.  However, data informing models are often temporally anchored (e.g. weekly case counts).  As a result, $R_0$ can be a difficult quantity to estimate and disentangle from time-based data.  For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-based parameters such as the beginning and end of an epidemic, two quantities that generally do not have precise definitions.} \ben{[Ben says: \@Shannon could we expand on this difficulty / show it really is difficult with 1 more sentence? It might benefit from listing the difficulting in estimating a robust $R_0$ value (difficulties: time steps used, scale and "start" and final start, assumptions about population growth and underlying model fit).]}\shannon{[Shannon says I added the few sentences to more smoothly introduce r0]} To demonstrate the difficulty of discerning $R_0$ in a time-dependent analysis, we first introduce \citet{Kermack1927}’s SIR model. This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,

@@

\ben{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$? Also, I feel like the story is weak here. The point is to leverage $R_0$ to show the value of time-invariant analysis - this seems a bit more like just discussing properties of $R_0$. In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}
\shannon{[Shannon says: Tried to tie this better to the previous part since it's no longer a new section.  also highlighted tie to TIA and r0.  I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}

\shannon{Time-invariant analysis helps practicioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations. Tne first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.  Both epidemics have populations of 1000 people with 10 individuals initially infected.  Additionally note that the two reproduction numbers are the same for each epidemic, $R_0 = 2=0.8/0.4 = 0.64/0.32$.  We plot the epidemics with traditional $state$ vs. $time$ plots. In Fig. \ref{fig:different-scales-standard} we show the time-based paths for the $S$, $I$, and $R$ states for the first 15 days of observed data.  In this time-variant view, we may believe that \ben{epidemic} 1 has a larger $R_0$ than \ben{epidemic} 2 because the peak of infection occurs more quickly than in Epidemic 2.  On the other hand, we may believe \ben{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \sout{model}\ben{epidemic} has not yet peaked at time 15.  In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$.

```{r echo = FALSE, fig.cap = "\\label{fig:different-scales-standard}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both epidemics are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$."  }
set.seed(1225)
theme_set(theme_bw(base_size = 24))

beta1 <- .8
gamma1 <- .4
scale <- .8
sir1 <- simulate_SIR_agents(n_sims = 100, n_time_steps = 40,
                            beta = beta1, gamma = gamma1,
                            init_SIR = c(990, 10, 0)) %>%
  group_by(sim) %>%
  agents_to_aggregate(states = c(tI, tR)) %>%
  group_by(t) %>%
  summarize(S = mean(X0), I = mean(X1), R = mean(X2),
            .groups = "drop") %>%
  mutate(type = "Epidemic 1")

sir2 <- simulate_SIR_agents(n_sims = 100, n_time_steps = 40,
                            beta = beta1 * scale, gamma = gamma1 * scale,
                            init_SIR = c(990, 10, 0)) %>%
  group_by(sim) %>%
    agents_to_aggregate(states = c(tI, tR)) %>%
  group_by(t) %>%
  summarize(S = mean(X0), I = mean(X1), R = mean(X2),
            .groups = "drop") %>%
  mutate(type = "Epidemic 2")

\shannon{NEW} A time-invariant approach to visualizing epidemics, in comparison, allows us to directly compare $R_0$ from a single plot.  In Fig. \ref{fig:different-scales-tern}
we plot the filamental trajectories of the two epidemics in a time-invariant view. We will explain how and why this works shortly. The important takeaway is that in this time-invariant view, it is apparent that these epidemics are on "the same path." In this case, this indicates that two epidemics have the same value of $R_0$.

\shannon{The reason why we can visually compare $R_0$ in Fig. \ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.} A filamental trajectory can be mathematically viewed as a set of points in space that have an ordering, and that all points on the line between these ordered points are also contained in the geometric object. For a SIR epidemic, we can represent the associated filamental trajectory $\psi$ as

$$
\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
$$
where a mapping $\xi : s \to \mathbb{R}$ that is monotonically increasing would not change the definition of $\psi$, i.e. $\psi_\xi = \psi$ where :

$$
\psi_\xi = \left \{(S(\xi(s)), I(\xi(s)), R(\xi(s))): S, I, R \ge 0, S + I + R = N \right \}_{s\in[0,T]}.
$$
Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then \ben{all points in} $\psi$ will lay in a  \ben{two-dimensional triangular plane in three-dimensional space.  \sout{We can then which can be visualize the \shannon{full filamental trajectory} in a two-dimensional ternary plot}  As a result, we can visualize \shannon{the full filamental trajectory} of an epidemic in a single \ben{ternary plot \sout{2d plot and ultimately $R_0$}}.  \shannon{[Shannon says: Show pic here?]}]\footnote{Which?  \shannon{3d space one? But I'm leaning against it now.   3d never looks good in a paper.}}

\sout{\ben{[This section could be a bit less wordy. But generally good.]} We visualize the two epidemics in a global, ternary view in Figure \ref{fig:different-scales-tern}.  Without getting into too much detail of the intricacies of this plot, we immediately see the points of the two filaments $\psi$ seem to form the same trajectory.  Now, it is much clearer that \ben{\sout{Model} epidemic} 2 is following the same trajectory as \ben{\sout{Model} epidemic} 1 but is not as far along in the infection process. }

The filamental trajectories in Fig. \ref{fig:different-scales-tern} seem to overlap, and we may suspect that something is fundamentally linking these two different epidemics together.  Mathematically, we can show this fundamental link turns is $R_0$.   Let our two epidemics be presented as $\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}$, $\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}$ respectively.  As with the example, assume both models have the same initial values $(S(0), I(0), R(0))$, and let $R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$. And define $a>0$ to be the relative scalar such that $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$.

@@

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.  The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model \ben{\sout{is also} can be mapped to} a point on the epidemic path of the second SIR model. \shannon{In other words, the two epidemics form the same filamental trajectory. For SIR models with similar initial state percentages, time-invariant analysis, time-invariant analysis allows practicioners to compare values of $R_0$ at a glance.}  \ben{[We've changed how we describe these two objects when they are related to the theorem - is that desirable? ]}\shannon{added that sentence to for that}

@@

@@

Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space allows us to better compare the overall structure of the epidemic and see how the population was directly impacted. \ben{\sout{In this section we will present} time-invariant tools \sout{that} can be useful even when the underlying generative model for the epidemic is unknown \shannon{or have more than three epidemic states}. \sout{We will also present how this tools can be used when the epidemics have more than 3 states.} \sout{trajectories, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states. }

\ben{New paragraph} Viewing epidemics as filamental trajectories provide \sout{a lot }new ways to compare and examine epidemics in a time-invariant manner. For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid \shannon{confusion stemming from}\sout{impacts of} temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point \shannon{are we missing a few words in this definition?}. This \shannon{representation induces}\sout{ also creates} a natural distance between this type of representation between epidemics, specifically:
\[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\psi_1'(s) - \psi_2'(s))^2 ds
\]
where $\psi_i'(s)$ the point along $\psi_1$ that is $s$ fraction of $|\psi_1|$ distance away from the start of $\psi_i$. This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall extremeness of points, including psuedo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}. This \shannon{what is this?} can be very useful when comparing between a set of simulation epidemics and the true epidemic, and does not constrain the number of states of the model, though we recommend projecting the points into the unit simplex\footnote{\ben{I'm not sure we've talked about this before... \shannon{I don't think we have but am wondering if we're getting in the weeds}}} (by making all values the proportion of the population in the given state).

\ben{New paragraph:} If the set of epidemics that one is examining have only gone through a single cycle of the outbreak (before the population of individuals have become susceptible again), then additional time-invariant tools, including prediction regions can we leveraged \shannon{awk sentence}^[{\shannon{What are we predicting if the epidemic is done?}}]. In these settings we go a step farther and treat epidemics more like geometric filaments \shannon{(i.e. filamental trajectories without an ordering of points)} than filamental trajectories. In \pkg{EpiCompare} we create prediction regions that  contain a the top ($1-\alpha$) proportion of simulated curves by defining geometric regions defined by the union of small filaments around the subset of simulations (subset by measures like the above pseudo-density estimates or depth estimates). These regions look at where in the state-space we expect the epidemic to traverse, and we can compare prediction regions defined by different models using the Haussdorf \shannon{why Haussdorf specifically?} distance as well as examining how well the truth epidemic matches the simulations by examining if the epidemic's trajectory lies within the prediction region. All these geometric structures and distances notations apply to epidemics with any number of states, and at the end of  Section \ref{sec:overview} we \sout{also} highlight how these prediction regions can aid in visual comparisons for epidemics with 3 states (like the SIR models).

@@

\shannon{I like your completion.  I think we can make it flow a bit more smoothly but the content we want to talk about is all there.}
